Predicting Survival Probability of NASA Aircraft Engines

### 2025-04-29 index.qmd
Final updates 4/29

```{r, echo = FALSE, warning=FALSE, include=FALSE}
#Load packages
library(tidyverse)
library(gridExtra)
library(grid)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)
library(sjPlot)
library(broom)

@@

As mentioned, CPH is used in various fields, primarily the health
industry, but the usage of the model does not come without challenges. A
COVID-19 study done using data from a Pakistani hospital noted that the
model was limited due to the recovery possibility of COVID-19 patients,
as CPH assumes as time goes on, the survival probability will approach
zero with no survivors. Though CPH was not the best model for this
study, the results of the model were that individuals who were
asymptomatic and young had a higher chance of surviving the virus
[@asghar2024]. In 2024, a study was done in the US on the relationship
between chest pain occurring and mortality of the person experiencing
chest pain using CPH. A response to that study demonstrated the various
limitations of CPH, with the primary concern being the constant hazards
ratio assumption that CPH relies on. In the heart study, the effect of
triglyceride-glucose was assumed to be a constant, but it is actually a
value that fluctuates over time along with an individual's body weight
and blood glucose levels. As such, the model would struggle to predict
correctly the effect of triglyceride-glucose given the violation of the
constant hazards ratio [@jiang2024].

An additional limitation of CPH comes from the data in the covariates
selected and how they can become biased. Two studies done using UK
health data demonstrated how bias can impact the values for the
covariates when relying on patient-submitted information. The first was
a study on the connection between salt intake and the development of
anxiety or depression after 14.5 years [@wang2025], and the second was a
study on the association between postpartum depression and the
development of two or more chronic diseases [@zhang2025]. Both of those
studies relied on patient-submitted health data and survey responses
supplemented with death records and other accessible health records.
This created a possible bias in their data that may have negatively
influenced their model as their collected data may not accurately
reflect the true data. These articles provide examples of the care that
needs to be taken when working with CPH models.

While health care is the dominant field in which CPH is used, a study in
2021 provides an example of using CPH in insurance fields. An insurance
company in the Czech Republic provides policies for certain illness
events, with the insured receiving money when the event occurs, which
concludes the contract. To assist the company with predicting if an
individual will have one of the illness events occur, a model using CPH
was generated using gender, age, and region as the covariates
[@zapletal2021]. A study in 2023 focused on the length of productive
life of female Floridian goats to determine the factors that would
impact the longevity of the goats. CPH was used to determine that both
the age at first birth and the specific herd the female goat belonged to
impacts the length of productive life [@ziadi2023]. Both of these
studies provide examples of the reaches of CPH models outside the health
industry.

@@

### Covariate Selection

Depending on the study, there will be covariates that do not impact the
probability of survival. Statistical methods such as the Wald test can
be used to determine which ones are statistically significant
individually and may have an impact in a multivariate model. Generating
a univariate CPH model for each covariate and using the Wald test with a
significance level of $\alpha = 0.05$ will allow for determination of if
a covariate is significant. Additional tests for models will involve
using a stepwise regression and comparing the results based on the
assumptions that are met and the concordance index. Any covariates that
are determined to be not significant to the survival probability can be
removed from the final CPH model.

@@

Based on the information provided by NASA on their data, there are
various sensors and operational setting fields that do not impact the
survival rate of an engine. To determine which covariates are
statistically significant, the below univariate test was run. The test
uses the Wald test to determine which covariates will significantly
impact the model individually. The covariates in which the p-value is
$< 0.05$ will be considered statistically significant.

```{r}
covariates <- c("os1", "os2", "os3", "sm1", "sm2", "sm3", "sm4", "sm5", "sm6", "sm7",
                "sm8", "sm9", "sm10", "sm11", "sm12", "sm13", "sm14", "sm15", "sm16",
                "sm17", "sm18", "sm19", "sm20", "sm21")

univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(time, status)~', x)))

univ_models <- lapply( univ_formulas, function(x){coxph(x, data = data_grouped)})

# Extract data
univ_results <- lapply(univ_models,
                       function(x){
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coefficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          HR <- paste0(HR, " (",
                                       HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test",
                                        "p.value")
                          return(res)
                         })
kable(as.data.frame(univ_results))
```

Based on the above test, none of the covariates are statistically
significant individually with all the data. As such, to determine the
best model to use for this analysis, a forward and backward stepwise
regression will be done to generate the models with the best Akaike
Information Criterion (AIC) and Bayesian Information Criterion (BIC)
values. This method produced three possible models. The fourth model in
the below table used covariates that were determined to significant with
the training data only.

The table below provides the model number, the covariates used, the AIC
and BIC from the stepwise regression if applicable, and the concordance
index. After determining the most accurate model, the CPH assumptions
will be tested.

@@

all_cov <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)

aic <- MASS::stepAIC(all_cov, trace = 0)
n <- nrow(data_grouped)
bic <- step(all_cov, direction = "both", k = log(n), trace = 0)

cox1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)

cox2 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm5 + sm8 + sm9 + sm15 + sm16 + sm17 + sm18 + sm20)

cox3 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm8 + sm9 + sm15 + sm18)

cox4 <- coxph(data = data_grouped, surv_object ~ os3 + sm13 + sm14 + sm19)

@@

```{r, echo = FALSE}
stepwise_results <- data.frame(c("model 1", "model 2", "model 3", "model 4"), 
                               c("All covariates", "os3, sm3 - sm5, sm8 - sm9, sm15 - sm18, sm20", "os3, sm3 - sm4, sm8-sm9, sm15, sm18", "os3, sm13 - sm14, sm19"),
                               c("2449.92", "2431.88", "N/A", "N/A"),
                               c("2547.71", "N/A", "2461.31", "N/A"),
                               c(cox1$concordance[6], cox2$concordance[6], cox3$concordance[6], cox4$concordance[6]))

@@

In determining the best model for this data, it was determined that CPH
is not the survival method that should be used on this data. By stepwise
regression, the models with the lowest BIC and AIC should have been used
in the following analysis; however, both models continuously failed the
CPH assumptions. As such, the model for the following analysis will be
model 1, which contained all covariates and had the highest concordance
value.

While model 1 does have initially contain all 24 covariates, there are
two that will need to be removed before proceeding with the analysis due
to infinite coefficients and linearly dependence. The covariate sm19 is
linearly dependent on another value, and the covariate sm16 resulted in
a possible infinite coefficient. The following analysis will be done
using model 1 without sm16 and sm19. The below tables provide the
summary values from the model.

@@

The forest plot for model 1 is shown below. The covariates that are
closest to the value of one have a hazard rate that has less effect on
the calculated hazard ratio for the entire model.

@@

While the concordance index of .696 achieved in this analysis demonstrates the model is fairly accurate, there is still room for improvement. Survival models resulting in a concordance index of .9 and higher are highly accurate and would provide a tighter, more accurate confidence interval for the longevity of engines. After attempting transformations of the data and creating models many different combinations of covariates, the final c-index that was calculated is the highest value achieved where the covariates were able to meet the proportional hazards model.

As discovered during the literature review, a weakness of the cox proportional hazards model is the necessity to meet the proportional hazards assumption. Due to the difficulty meeting the assumption during this analysis, to achieve a more accurate model it would be beneficial to explore other survival models for this dataset.

>>>>

```{r, echo = FALSE, warning=FALSE, include=FALSE}
#Load packages
library(tidyverse)
library(gridExtra)
library(grid)
library(survival)
library(survminer)
library(ggpubr)
library(magrittr)
library(ggfortify)
library(knitr)
library(sjPlot)
library(broom)
library(glmnet)

@@

As mentioned, CPH is used in various fields, primarily the health
industry, but the usage of the model does not come without challenges. A
COVID-19 study using data from a Pakistani hospital noted that the model
was limited due to the recovery possibility of COVID-19 patients. CPH
assumes as time goes on, the survival probability will approach zero
with no survivors, which was not the case with COVID-19. Though CPH was
not the best model for this study, the results of the model were that
individuals who were asymptomatic and young had a higher chance of
surviving the virus [@asghar2024]. In 2024, a study was done in the US
on the relationship between chest pain occurring and mortality of the
person experiencing chest pain using CPH. A response to that study
demonstrated the various limitations of CPH, with the primary concern
being the constant hazards ratio assumption that CPH relies on. In the
heart study, the effect of triglyceride-glucose was assumed to be a
constant, but it is actually a value that fluctuates over time along
with an individual's body weight and blood glucose levels. As such, the
model would struggle to predict correctly the effect of
triglyceride-glucose given the violation of the constant hazards ratio
[@jiang2024].

An additional limitation of CPH comes from the data as the selection of
covariates and the generation of the data can be biased. Two studies
done using UK health data demonstrated how bias can impact the values
for the covariates when relying on patient-submitted information. The
first was a study on the connection between salt intake and the
development of anxiety or depression after 14.5 years [@wang2025], and
the second was a study on the association between postpartum depression
and the development of two or more chronic diseases [@zhang2025]. Both
of those studies relied on patient-submitted health data and survey
responses supplemented with death records and other accessible health
records. The reliance This created a possible bias in their data that
may have negatively influenced their model as their collected data may
not accurately reflect the true data. These articles provide examples of
the care that needs to be taken when working with CPH models.

While health care is the dominant field in which CPH is used, a study in
2021 provides an example of using CPH in insurance fields. An insurance
company in the Czech Republic provides policies for certain illness
events, with the insured receiving money when the event occurs, which
concludes the contract. To assist the company with predicting if an
individual will have one of the illness events occur, a model using CPH
was generated using gender, age, and region as the covariates
[@zapletal2021]. A study in 2023 focused on the length of productive
life of female Floridian goats to determine the factors that would
impact the longevity of the goats. CPH was used to determine that both
the age at first birth and the specific herd the female goat belonged to
impacts the length of productive life [@ziadi2023]. Both of these
studies provide examples of CPH models outside the health industry.

@@

@@

Based on the information provided by NASA on their data, there are
various sensors and operational setting fields that do not impact the
survival rate of an engine. To determine which covariates are
statistically significant, the below univariate test was run. The test
uses the Wald test to determine which covariates will significantly
impact the model individually. The covariates in which the p-value is
$< 0.05$ will be considered statistically significant. The table
displays the covariate, the $\beta$ value, the hazard ratio (HR), the
95% confidence interval for the hazard ratio, and the p-value generated
from the Wald test. As can be seen, none of the covariates are
statistically signification on their own.

@@

covariates <- c("os1", "os2", "os3", "sm1", "sm2", "sm3", "sm4", "sm5", "sm6", "sm7",
                "sm8", "sm9", "sm10", "sm11", "sm12", "sm13", "sm14", "sm15", "sm16",
                "sm17", "sm18", "sm19", "sm20", "sm21")

forms <- sapply(covariates,
                        function(x) as.formula(paste('Surv(time, status)~', x)))

univ_models <- lapply(forms, function(x){coxph(x, data = data_grouped)})

# Extract data
univ_results <- lapply(univ_models,
                       function(x){
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coefficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          CI <- paste0("(", HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, CI, p.value)
                          names(res)<-c("beta", "HR", "95% CI", 
                                        "p-value")
                          return(res)
                         })
kable(as.data.frame(univ_results))
```

To determine the best model to use for this analysis, a forward and
backward stepwise regression was done to generate the models with the
best Akaike Information Criterion (AIC) and Bayesian Information
Criterion (BIC) values. The second and third models were generated using
this regression method. The fourth model was generated using the Lasso
penalized Cox regression method.

The table below provides the model number, the covariates used, the AIC
and BIC from the stepwise regression if applicable, and the concordance
index.

```{r}
all_cov <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)

#stepwise regression models
aic <- MASS::stepAIC(all_cov, trace = 0)
n <- nrow(data_grouped)
bic <- step(all_cov, direction = "both", k = log(n), trace = 0)

#input/out for glmnet
all_mod <- model.matrix(Surv(time, status) ~  + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21, data = data_grouped)[, -1]

# lasso penalized cox
cvfit <- cv.glmnet(all_mod, surv_object, family = "cox")
coefs <- coef(cvfit, s = "lambda.min")
new_cov <- rownames(coefs)[which(coefs != 0)]

#all Cox models 
#all covariates
cox1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)

#aic
cox2 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm5 + sm8 + sm9 + sm15 + sm16 + sm17 + sm18 + sm20)

#bic
cox3 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm8 + sm9 + sm15 + sm18)

#lasso
cox4 <- coxph(data = data_grouped, surv_object ~ os2 + sm1 + sm3 + sm4 + sm9 + sm10 + sm11 + sm14 + sm15 + sm16 + sm17 + sm18)

@@

```{r, echo = FALSE}
stepwise_results <- data.frame(c("model 1", "model 2", "model 3", "model 4"), 
                               c("All covariates", "os3, sm3, sm4, sm5, sm8, sm9, sm15, sm16, sm17, sm18, sm20", "os3, sm3, sm4, sm8, sm9, sm15, sm18", "os2, sm1, sm3, sm4, sm9, sm10, sm11, sm14, sm15, sm16, sm17, sm18"),
                               c("2449.92", "2431.88", "N/A", "N/A"),
                               c("2547.71", "N/A", "2461.31", "N/A"),
                               c(cox1$concordance[6], cox2$concordance[6], cox3$concordance[6], cox4$concordance[6]))

@@

All four models were checked for the CPH assumptions. All but the first
model failed the assumptions, even after adapting the covariate using
stratification and/or removal. As such, the first model was the one
selected for the continuing analyses.

While model one does have initially contain all 24 covariates, there are
two that will need to be removed before proceeding with the analysis due
to infinite coefficients and linearly dependence. The covariate sm19 is
linearly dependent on another value, and the covariate sm16 resulted in
a possible infinite coefficient. The following analysis will be done
using model one without sm16 and sm19. The below tables provide the
summary values from the model.

@@

The forest plot for model one is shown below. The covariates that are
closest to the value of one have a hazard rate that has less effect on
the calculated hazard ratio for the entire model.

@@

While the concordance index of .696 achieved in this analysis
demonstrates the model is fairly accurate, there is still room for
improvement. Survival models resulting in a concordance index of .9 or
higher are highly accurate and would provide a tighter, more accurate
confidence interval for the longevity of engines. After attempting
transformations of the data and creating models with many different
combinations of covariates, the final c-index that was calculated is the
highest value achieved where the covariates were able to meet the
proportional hazards model.

As discussed during earlier, a weakness of the Cox proportional hazards
model is the necessity to meet the proportional hazards assumption.
Given the difficulty of meeting the PH assumption, a different survival
analysis would have better suited this dataset.
